---
title: "ST558 Project 1"
author: "Zack Vaskalis"
date: "6/12/2020"
output: 
  rmarkdown::github_document:
    toc: TRUE
    toc_depth: 3
    html_preview: TRUE
urlcolor: blue
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
```

# Reading and Summarizing JSON Datasets 

A Vignette using National Hockey League (NHL) Application Programming Interface (API) - No "Hat Tricks" here!   

```{r jsonImage, out.width = "200px", echo=FALSE}
knitr::include_graphics("https://upload.wikimedia.org/wikipedia/commons/c/c9/JSON_vector_logo.svg")
```

## The power of JSON

### What is JSON?

Before we can begin investigating the datasets provided by the [NHL Records 'Franchise' API](https://gitlab.com/dword4/nhlapi/-/blob/master/records-api.md), we must first discuss what JSON is and what tools do we have at our disposal in R to handle reading such data into R.

JSON stands for **JavaScript Object Notation** and is a language-independent data format, derived from _JavaScript_, but is an open standard file and data interchange format used to store and transmit data.  Simply put, JSON files are plain text that are readable by humans and can be easily opened and inspected. JSON data files use the extension `.json` and encode attribute-value pairs, such as `"firstName": "Zachary"`, simple array data types such as `"sentence": ["Hello", "World", "I", "am", "Zack"]`, to more complicated data structures or objects that can be stored (in a file or buffer), transmitted (via network connection) and reconstructed later, back into a more complicated structure, like a matrix, data frame, or tibble, potentially in a completely different environment.  JSON was developed in part to solve the problem for the need of a real-time server-to-browser communication protocol which was independent of popular browser-plugins applets like Flash and Java, which were popular at the start of the new millenium.  This is exactly why JSON has become so popular with widespead use - virtually every programming language can process and deal with plain text data, and the files are easy to transfer over the internet.

### JSON Data Types

JSON has 5 basic data types: numbers, strings, boolean or logical (including the empty value, `NULL`), arrays, and objects.  These 5 basic JSON data types, adapted from the [JSON wikipedia page](https://en.wikipedia.org/wiki/JSON) are outlined below.  Whitespace special characters such as spaces, horizontal tabs, newlines, and returns are allowed and typically ignored outside of values and punctuation marks, but are maintained when contained within string values for instance.   

*  Number:  a signed decimal number (that can consist of a fractional component and/or exponential component), but cannot be a non-number.
*  String:  a sequence of Unicode characters, delimited by double-quotation marks, and supports escaping syntax utilizing a backslash. 
*  Boolean: (or Logical) dichotomous data type or flag consisting of one of two values: `TRUE` or `FALSE`.  Also allows for empty value `NULL`.
*  Array:   an ordered list of values where each one may be a different type, delimited using square bracket notation with elements separated by commas.
*  Object:  a collection of name–value string pairs, delimited by curly brackets and separated by commas.  Within each pair, a colon character separates the name from its value.

### So where is JSON used?

A great little quick guide tutorial and overview on JSON can be found [here](https://www.tutorialspoint.com/json/json_quick_guide.htm#:~:text=JSON%20format%20is%20used%20for,used%20with%20modern%20programming%20languages.) if you are interested in a more extensive dive with some examples and applications using JSON in several programming languages.  There are several areas where JSON is used, however, we've already discussed the main one which is primarily for used for serializing and transmitting structured plain-text data over a network connection, transmitting this data between a server and various web applications, serving as an alternative to, and in some regards seems to be slowly replacing
Extensible Markup Language (XML).  For more on this somewhat heated debate, check out these articles: 

*  [XML is toast, long live JSON](https://www.cio.com/article/3082084/xml-is-toast-long-live-json.html)
*  [JSON vs XML](https://hackr.io/blog/json-vs-xml)
*  [Has JSON Overtaken XML](https://www.c-sharpcorner.com/article/is-json-overridden-xml/#:~:text=XML%20has%20done%20a%20lot,still%20exist%20in%20this%20world.)
*  [Is XML Dying?](https://blog.submain.com/json-and-xml-dying/)

Tied to the reason above, web services and Application Programming Interfaces (APIs) use JSON format to provide public data.  Thus, it is used nearly everywhere, in particular in areas which were previously troubling such as with cross-domain communication.  As mentioned previously, with the dawn of JavaScript and XML, AJAX (Asynchronous JavaScript and XML), it became of greater importance for websites to be able to load data sources quickly, asynchronously, or in the background, all without slowing down the speed of loading and rendering the page.  Modern sites, like for example Twitter, use RSS (**R**esource Description Framework **S**ite **S**ummary) feeds, which is essentially XML-formatted plain text originally designed as a metadata data model.  Before we get too far down the proverbial rabbit hole, RSS feeds became popular because on the server-side they are easy to use and import.  However, if you try to use AJAX on them, you will get errors, because they are only able to load in the same domain the request is being made from - generally querys will result in errors.  This is truly where JSON shines, overcoming this inability to communicate across domains, and therefore is able to send data packets back to the domain making the call or query.  This is why we will be using JSON to query the NHL Hockey API without worry.     

### A good way to store data?

For all of the reasons we have discussed above, JSON is a great way to store data because it is light-weight, easy to read and write, plain-text based, highly interoperable since it is program language independent, and overcomes the cross-domain communication barrier - where JSON files are easily sent between servers. 
This may not be true for all data - especially extremely private data like social security numbers and income records for example.  Although JSON would be able to handle the data, you would probably want this type of data stored on more securely encrypted servers, which are not easily accessed.  But public data, especially governmental data, national health organization data, or even as in our example here, major league sports team data, JSON is completely appropriate and easy to use, besides being a great way to store data.

## Possible R Packages for JSON

So now that we have come to a basic level of understanding what JSON data is, where it is used, and why it is a good way to store data, we need to learn how to deal with JSON data.  This portion of the discussion is highly programming language specific, and in our case for the rest of this vignette, the R programming language will be the method of choice. Thus, to handle JSON data in R, it is necessary to discuss the packages and libraries, along with their associated functions, which were designed and made available for reading in JSON data in R.  There are several major packages for handling JSON data in R, which are as follows:

*  `rjson`
*  `RJSONIO`
*  `jsonlite`
*  `tidyjson`

There is a great resource, [A biased comparsion of JSON packages in R](http://rstudio-pubs-static.s3.amazonaws.com/31702_9c22e3d1a0c44968a4a1f9656f1800ab.html), which compares the top three packages listed above, and is in favor of `jsonlite`, which I would have to agree with, out of that list jsonlite is by far the best, for some of the reasons mentioned in that resource.  

However, published on the [CRAN R Project website](https://cran.r-project.org/) a little under two weeks ago, is a new package `tidyjson` which seems to have some promise.  What is especially intriguing is the writeup on the `tidyjson` package page itself which can be found on the  [CRAN R Project website for tidyjson](https://cran.r-project.org/web/packages/tidyjson/vignettes/introduction-to-tidyjson.html#:~:text=Tidyjson%20provides%20a%20grammar%20for,work%20with%20in%20the%20tidyverse.)  I am presenting it here so that you know it is available as a resource.  With it being so new, I believe it is best to utilize `jsonlite`.

The resource mentioned previously, [A biased comparsion of JSON packages in R](http://rstudio-pubs-static.s3.amazonaws.com/31702_9c22e3d1a0c44968a4a1f9656f1800ab.html) describes many benefits for using `jsonlite` to deal with JSON data that I agree with, so I will not rehash those here.  Additionally, however, besides `jsonlite` being the package we've explored so far in this course, I believe there are other benefits.  From the [CRAN R webiste for jsonlite](https://cran.r-project.org/web/packages/jsonlite/index.html) we see that `jsonlite` was originally built as a fork of `RJSONIO` but has since been completely rewritten.  The site also claims that `jsonlite` is a fast, robust, high-performance JSON parser and generator for R, optimized for statistical data and the web, and offers flexible, robust, high performance tools for working with JSON in R and is particularly powerful for building pipelines and interacting with a web API.  These are exactly all the things we want to do as part of this vignette, so `jsonlite` seems like the perfect choice.  Finally, `jsonlite`, in addition to converting JSON data to R objects, contains functions that can stream, validate, and prettify JSON data.  Again, this is what we want to do, so why not use it?!?

## NHL API JSON Data

Now that we know all about JSON data and the R packages, namely `jsonlite` that can help us deal with reading in such JSON data and creating R objects, we can now begin to explore the NHL data provided by the [NHL Records 'Franchise' API](https://gitlab.com/dword4/nhlapi/-/blob/master/records-api.md).

We will use R to build 5 functions that return well-formatted, parsed data from this API.  The code chunks we develop will use the following NHL Records Franchise API calls:

1.  /franchise (Returns id, firstSeasonId and lastSeasonId and name of every team in the history of the NHL)
2.  /franchise-team-totals (Returns Total stats for every franchise (ex roadTies, roadWins, etc))
3.  /franchise-season-records?cayenneExp=franchiseId=ID (Drill-down into season records for a user-specified franchise ID)
4.  /franchise-goalie-records?cayenneExp=franchiseId=ID (Goalie records for a user-specified franchise ID)
5.  /franchise-skater-records?cayenneExp=franchiseId=ID (Skater records, for a user-specified franchise ID)

### 1. NHL Franchises

Let's start with number 1.  Function `getFranchises()` will query the NHL Franchise API to return a data table with all of the NHL franchises, what the team place name and common name are, what the team and franchise IDs are for ALL of the JSON data being queried, as well as the year of their first season (written as 19171918 to indicate it is the 1917-1918 season), as well as the lastSeasonId if that specific franchise no longer exists.    

```{r LoadLibraries, include=FALSE, echo=TRUE}
#initialize all needed libraries
library(tidyverse)
library(pipeR)
library(DT)
library(dplyr)
library(jsonlite)
library(httr)
library(ggplot2)
```

```{r franchise, echo=TRUE}
getFranchises <- function(...){
#franchise json data.  Set up components of URL for API query using GET() function
  query_prefix <- 'https://records.nhl.com/site/api'
  query_suffix <- '/franchise'
  franchiseJSON <- GET(paste0(query_prefix,query_suffix))

#Transform Data from JSON to a data frame
  franchiseData <- fromJSON(content(franchiseJSON,"text", encoding = "UTF-8"),flatten = TRUE) 

#Rename id to franchiseID
  franchiseData <- rename(franchiseData$data, franchiseID = id)
  franchiseData

#Reorganize the data frame with columns in a more usable order
  franchiseData %>% select(mostRecentTeamId, franchiseID, teamPlaceName, teamCommonName, firstSeasonId, lastSeasonId) %>%
    arrange(teamPlaceName) %>>% (~ franchiseData)
  franchiseDT <- datatable(franchiseData, rownames = FALSE)

#Initialize franchiseSelection Table for user to be able to see correct franchise ID for functions #3 - 5.
  franchiseSelection <- franchiseData

#return datatable
  return(list(franchiseDT,invisible(franchiseSelection)))
}
```

```{r callgetFranchise, echo=TRUE}
getFranchises()[[1]]
```


